### 2022-07-19 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@23efa8475c06ad8174f1be4d1e9041c517d5daf8 🚀

The principle of maximum entropy (MaxEnt) is a statistical and
topological model that can be used for both of these purposes, i.e.~to
make predictions of network structure and to better understand processes
shaping ecological networks. This mathematical method, briefly presented
in Box 1, has been used in a wide range of disciplines, from
thermodynamics to chemistry and biology (Martyushev and Seleznev 2006).
It has also been proven useful in ecology, e.g.~in species distribution
models (Phillips, Anderson, and Schapire 2006) and macroecological
models (Harte et al. 2008; Harte and Newman 2014). As discussed in Box
1, maximizing a measure of entropy ensures that the derived probability
distributions are unique and least biased under the set of constraints
used. These constraints are built using state variables, i.e.~variables
that represent the macrostate of the system. The challenge is to find
the set of state variables that best represent natural systems and to
translate them into appropriate statistical constraints. Having a
validated maximum entropy model for the system at hand allows us to make
rigorous predictions using a minimal amount of data, as well as helping
us describe the most important factors driving that system.

@@

In this contribution, we used two complementary approaches to predict
the structure of food webs using the principle of maximum entropy. We
then compared our predictions against empirical data and null and
neutral models commonly used in network ecology. The first approach
consists in deriving constrained probability distributions of given
network properties directly. We derived the joint degree distribution (a
probability distribution) of maximum entropy using only the number of
species \(S\) and the number of interactions \(L\) as state variables.
Then, we predicted the degree distribution of maximum entropy directly
from the joint degree distribution since the first is the sum of the
marginal distributions of the second (a species' degree is the sum of
its in and out-degrees). Because of the scarcity of empirical data on
the number of interactions in ecological networks, we present a method
to predict \(L\) from \(S\) (Box 2), thus allowing the prediction of the
joint degree distribution from \(S\) solely. In turn, the second
approach consists in finding, using different constraints, the adjacency
matrix of maximum entropy from which network properties can be measured.
To do so, we used a flexible and heuristic approach based on simulated
annealing to find networks \emph{close} to maximum entropy. As discussed
above, our choice of algorithm stands from the very nature of food webs
(i.e.~simple directed networks allowing self-loops) that makes the
analytical derivation of a maximum entropy graph model difficult. We
first built our type I MaxEnt network model constrained by the
connectance of the network (i.e.~the ratio \(L/S^2\)). A comparison of
this model against empirical data indicated that connectance alone was
not sufficient to predict many aspects of network structure. For this
reason, we built our type II MaxEnt network model, which instead uses
the whole joint degree sequence as a constraint. Overall, we found that
this second model was much better at predicting food-web structure than
the one based on connectance.

\hypertarget{box-1---the-principle-of-maximum-entropy-a-primer-for-ecologists}{%
\section{Box 1 - The principle of maximum entropy: A primer for
ecologists}\label{box-1---the-principle-of-maximum-entropy-a-primer-for-ecologists}}

@@

In this contribution, we also use the SVD entropy as a measure of
entropy, which is an application of Shannon's entropy to the relative
non-zero singular values of a truncated singular value decomposition
(t-SVD; Strydom, Dalla Riva, and Poisot 2021) of a food web's Boolean
adjacency matrix. This measure also satisfies all four properties
above-mentioned, while being a proper measure of the internal complexity
of food webs (Strydom, Dalla Riva, and Poisot 2021). We measured SVD
entropy as follows:

@@

where \(s_i\) are the relative singular values
(\(s_i = \sigma_i / \sum_{i = 1}^R \sigma_i\), where \(\sigma_i\) are
the singular values). Following Strydom, Dalla Riva, and Poisot (2021),
we standardized this measure with the rank \(R\) of the matrix
(i.e.~\(J / \ln(R)\)) to account for the difference in dimensions
between networks (\emph{sensu} Pielou's evenness; Pielou 1975). In a
following section, we will show how SVD entropy can be used to predict a
network of maximum entropy (i.e.~of maximum complexity) heuristically.

\hypertarget{testing-maxent-models}{%
\section{Testing MaxEnt models}\label{testing-maxent-models}}

\hypertarget{data}{%
\subsection{Data}\label{data}}

@@

where \(\lambda_1\), \(\lambda_2\), and \(\lambda_3\) are the Lagrange
multipliers. The probability distribution that maximizes entropy is
obtained by finding these values. As pointed out in Box 1, \(F\) is just
Shannon's entropy to which we added terms that each sums to zero (our
constraints). \(F\) is maximized by setting to 0 its partial derivative
with respect to \(p(k_{in},k_{out})\). Because the derivative of a
constant is zero, this gives us:

@@

In fig.~S2, we show that the degree distribution of maximum entropy,
given \(S\) and \(L\), predicts very low probabilities that a species
will be isolated in its food web (i.e.~having \(k = 0\)). As MacDonald,
Banville, and Poisot (2020) pointed out, the size of food webs should at
least be of \(S-1\) interactions, since a lower number would yield
isolated species, i.e.~species without any predators or preys. Our
results show that, under our purely information-theoretic model, the
probability that a species is isolated is quite high below this
threshold. The expected proportion of isolated species rapidly declines
by orders of magnitude with increasing numbers of species and
interactions.

@@

\hypertarget{box-2---working-with-predicted-numbers-of-interactions}{%
\section{Box 2 - Working with predicted numbers of
interactions}\label{box-2---working-with-predicted-numbers-of-interactions}}

@@

@@

We define networks of maximum entropy as the configuration of the
adjacency matrix with the highest SVD entropy under a set of
constraints. As mentioned in Box 1, we used the SVD entropy as our
measure of entropy since it has been shown to be a reliable measure of
food-web complexity (Strydom, Dalla Riva, and Poisot 2021), in addition
to having the required properties of a proper measure of information
entropy. We thus seek to find the network with the highest complexity,
or randomness, that exactly reproduces specified constraints on its
structure. Our method is in contrast with maximum entropy graph models
that predict a probability distribution on networks under soft or hard
constraints (e.g., Park and Newman 2004; Cimini et al. 2019). We believe
our approach to be more flexible, easier to compute, while allowing
direct comparisons of empirical food webs with more complex networks
with similar structure.

>>>>

The principle of maximum entropy (MaxEnt) is a statistical and
topological model that can be used for both of these purposes, i.e.~to
make predictions of network structure and to better understand processes
shaping ecological networks. Food-web properties that can be derived
using MaxEnt are varied and pertain to different elements of the network
(i.e.~at the species (node), the interaction (edge) or the community
(network) levels). For example, at the species level, MaxEnt can be used
to predict the distribution of trophic levels among species, as well as
the distribution of species' vulnerability (number of predators) and
generality (number of preys). By contrast, at the interaction level,
predictions can be made on the distribution of interaction strengths in
weighted food webs. At the community level, it can give us a probability
distribution of potential networks or of specific measures of their
emerging structure. The applicability and performance of this method
depend on the ecological information available and on our capacity to
find the set of state variables that best represent natural systems and
to translate them into appropriate statistical constraints. MaxEnt has
been used in a wide range of disciplines, from thermodynamics to
chemistry and biology (Martyushev and Seleznev 2006). It has also been
proven useful in ecology, e.g.~in species distribution models (Phillips,
Anderson, and Schapire 2006) and macroecological models (Harte et al.
2008; Harte and Newman 2014). Maximizing a measure of entropy ensures
that the derived probability distributions are unique and least biased
under the set of constraints used. These constraints are built using
state variables, i.e.~variables that represent the macrostate of the
system. Having a validated maximum entropy model for the system at hand
allows us to make rigorous predictions using a minimal amount of data,
as well as helping us describe the most important factors driving that
system.

@@

In this contribution, we used two complementary approaches to predict
the structure of food webs using the principle of maximum entropy. We
then compared our predictions against empirical data and null and
neutral models commonly used in network ecology. The first approach
consists in deriving constrained probability distributions of given
network properties directly. We derived the joint degree distribution (a
probability distribution) of maximum entropy using only the number of
species \(S\) and the number of interactions \(L\) as state variables.
Then, we predicted the degree distribution of maximum entropy directly
from the joint degree distribution since the first is the sum of the
marginal distributions of the second (a species' degree is the sum of
its in and out-degrees). Because of the scarcity of empirical data on
the number of interactions in ecological networks, we present a method
to predict \(L\) from \(S\) (Box 1), thus allowing the prediction of the
joint degree distribution from \(S\) solely. In turn, the second
approach consists in finding, using different constraints, the adjacency
matrix of maximum entropy from which network properties can be measured.
To do so, we used a flexible and heuristic approach based on simulated
annealing to find networks \emph{close} to maximum entropy. As discussed
above, our choice of algorithm stands from the very nature of food webs
(i.e.~simple directed networks allowing self-loops) that makes the
analytical derivation of a maximum entropy graph model difficult. We
first built our type I MaxEnt network model constrained by the
connectance of the network (i.e.~the ratio \(L/S^2\)). A comparison of
this model against empirical data indicated that connectance alone was
not sufficient to predict many aspects of network structure. For this
reason, we built our type II MaxEnt network model, which instead uses
the whole joint degree sequence as a constraint. Overall, we found that
this second model was much better at predicting food-web structure than
the one based on connectance.

\hypertarget{the-principle-of-maximum-entropy-a-primer-for-ecologists}{%
\subsection{The principle of maximum entropy: A primer for
ecologists}\label{the-principle-of-maximum-entropy-a-primer-for-ecologists}}

@@

Shannon's entropy can only be calculated on conventional probability
distributions such as the joint degree distribution. This is an issue
when working with ecological networks, which are frequently represented
as adjacency matrices. For this reason, in this contribution, we also
use the SVD entropy as a measure of entropy, which is an application of
Shannon's entropy to the relative non-zero singular values of a
truncated singular value decomposition (t-SVD; Strydom, Dalla Riva, and
Poisot 2021) of a food web's Boolean adjacency matrix. We measured SVD
entropy as follows:

@@

where \(s_i\) are the relative singular values of the adjacency matrix
(\(s_i = \sigma_i / \sum_{i = 1}^R \sigma_i\), where \(\sigma_i\) are
the singular values). Note that the distribution of relative singular
values is analogous to a probability distribution, with \(0 < s_i < 1\)
and \(\sum s_i = 1\). Thus, this measure also satisfies all four
properties above-mentioned, while being a proper measure of the internal
complexity of food webs (Strydom, Dalla Riva, and Poisot 2021).
Following Strydom, Dalla Riva, and Poisot (2021), we standardized this
measure with the rank \(R\) of the matrix (i.e.~\(J / \ln(R)\)) to
account for the difference in dimensions between networks (\emph{sensu}
Pielou's evenness; Pielou 1975). We will show how SVD entropy can be
used to predict the network of maximum entropy (i.e.~of maximum
complexity) heuristically given a predetermined set of constraints.

\hypertarget{data-and-code}{%
\section{Data and code}\label{data-and-code}}

@@

where \(\lambda_1\), \(\lambda_2\), and \(\lambda_3\) are the Lagrange
multipliers. The probability distribution that maximizes entropy is
obtained by finding these values. As pointed out in the introduction,
\(F\) is just Shannon's entropy to which we added terms that each sums
to zero (our constraints). \(F\) is maximized by setting to 0 its
partial derivative with respect to \(p(k_{in},k_{out})\). Because the
derivative of a constant is zero, this gives us:

@@

In fig.~S2, we show that the degree distribution of maximum entropy,
given \(S\) and \(L\), predicts very low probabilities that a species
will be isolated in its food web (i.e.~having \(k = 0\)). As MacDonald,
Banville, and Poisot (2020) pointed out, the size of food webs should at
least be of \(S-1\) interactions, since a lower number would yield
isolated species, i.e.~species without any predators or preys. Our
results show that, under our purely information-theoretic model, the
probability that a species is isolated is quite high when the total
number of interactions is below this threshold. The expected proportion
of isolated species rapidly declines by orders of magnitude with
increasing numbers of species and interactions.

@@

\hypertarget{box-1---working-with-predicted-numbers-of-interactions}{%
\section{Box 1 - Working with predicted numbers of
interactions}\label{box-1---working-with-predicted-numbers-of-interactions}}

@@

\hypertarget{box-2---corresponding-null-and-neutral-models}{%
\section{Box 2 - Corresponding null and neutral
models}\label{box-2---corresponding-null-and-neutral-models}}

@@

We define networks of maximum entropy as the configuration of the
adjacency matrix with the highest SVD entropy under a set of
constraints. As mentioned above, we used the SVD entropy as our measure
of entropy since it has been shown to be a reliable measure of food-web
complexity (Strydom, Dalla Riva, and Poisot 2021), in addition to having
the required properties of a proper measure of information entropy. We
thus seek to find the network with the highest complexity, or
randomness, that exactly reproduces specified constraints on its
structure. Our method is in contrast with maximum entropy graph models
that predict a probability distribution on networks under soft or hard
constraints (e.g., Park and Newman 2004; Cimini et al. 2019). We believe
our approach to be more flexible, easier to compute, while allowing
direct comparisons of empirical food webs with more complex networks
with similar structure.
